---
plottitle: "Initial data clean-up: fisheries data"
output: word_document
---

Some initial summarization and plotting of data can often reveal errors. In this example, we continue with some of the WDFW fisheries data for 2014. 

```{r}
library(tidyr)
library(dplyr)
library(ggplot2)
df <- read.csv("D:/fishdata.csv", header = TRUE, sep = ",")
```

Initially, we create an xy plot of Weight x Length for the entire dataset and including all species. 


```{r, echo=TRUE}
qplot(Length_meas, Weight_meas, data = df)
```

It's clear that there are some outliers at the lengths <200. So we'll filter the data and replot for these smaller lengths. 

```{r, echo=TRUE}
df200 <- df %>%
  filter(Length_meas < 200)
qplot(Length_meas, Weight_meas, data = df200)
```

This chart makes it clear that there are 3 unusual observations. Often these are due to clerical errors (e.g., decimal in the wrong place). 

We can again filter the dataset to extract the outlying observations.

```{r, echo=TRUE}
dfoutliers <- df %>%
  filter(Length_meas < 200) %>%
  filter(Weight_meas > 250)
dfoutliers
```

The outlying points can either be corrected or removed from the dataset as appropriate.There may be additional outliers in the data for measurements >200, but this won't be clear until the data are separated by species.

```{r, echo=TRUE}
library(lattice)
species.f <- factor(df$Species_code)
xyplot(df$Weight_meas~df$Length_meas|species.f)
```




