---
title: "Tidying and using data"
output:
  ioslides_presentation:
    highlight: tango
    incremental: yes
  beamer_presentation: null
---

##What is a workflow?
*A series of steps necessary to complete a task*

For example, I hate flight delays! I want to avoid days and times that have a lot of delays.

To minimize my risk of flight delays, I should take a large dataset of flights, 

then select those flights that had delays,

then group those flights by date and hour, 

then summarize by mean flight delay,

then select those times with delays over 2 hours

*Then avoid traveling on those days and times!*

##dplyr allows me to do just that with the "pipeline"

x %>% f(y) -> f(x, y)

hourly_delay <- flightdata %>%
  filter(!is.na(dep_delay)) %>%
  group_by(date, hour) %>%
  summarise(
    delay = mean(dep_delay),
n = n() ) %>%
  filter(n > 2)
  
  
##dplyr

- **filter:** keep rows that match a criteria

- **select**: pick columns by name

- **arrange:** reorder rows

- **mutate:** add new variables

- **summarize:** reduce variables to values

- **+ group_by**


##What are tidy data?

Data that are easy to transform, visualize and model

Key idea: Store variables in a consistent way:

- Each variable forms a column
- Each observation forms a row
- Each type of observational unit forms a table

##tidyr: 

- **Gather:** convert many columns into variable/value pairs; akin to melt in reshape

- **Spread:** convert variable/value pairs into columns; akin to cast in reshape

- **Separate:** break one column into two (similar to strsplit)

##The five most common problems with messy data:

- Multiple types of observational units are stored in the same table
- Column headers are values, not variable names
- Multiple variables are stored in one column
- Variables are stored in both rows and columns
- A single observational unit is stored in multiple tables

##Let's develop a workflow to clean some messy data
First we have to load the packages
```{r, warning=FALSE, message=F}
library(dplyr)
library(tidyr)
library(knitr)
```


##Let's have a look at that messy data
```{r}
rawdat<-read.csv("Niwot_Salix_2014_WillowSeedlingSurvey.csv", fill=T)
kable(head(rawdat))
```

##Multiple observational units in the same table

- The meta data in is the table!
- Includes replication at both the individual level and the plot level
- Code, snow, n and temp are repeated (or in this case worse - not repeated!)

##Multiple units in same table fix 1: remove the meta data

First we need to assign correct column names
```{r}
mycolnames<-t(rawdat[10,])
colnames(rawdat)<-c(mycolnames)
```

Then delete the metadata rows

```{r}
dat<-rawdat[-c(1:10), ]
```

##What's that look like?
```{r}
kable(head(dat))
```

##Multiple units in same table fix 2: separate the treatment variables into their own table
```{r}
#create a vector of equal to plot number repeated 5 times each
myhack<-rep(1:48, each=5)
#rbind that with dat
dat2<-cbind(myhack, dat)
```


##Compile the plot-level data
```{r}
#select the treatment variables
mytrts<-dat2[c("myhack", "block", "plot", "code", "snow", "n", "temp")]
#subset that to the unique values
treatment_dat<-subset(mytrts, mytrts$block!="")
#remove the hack
treatment_dat$myhack<-NULL
```

##What's that look like?
```{r}
kable(head(treatment_dat))
```

##Compile the individual-level data
```{r}
#select the individual columns
myindivs<-c(1, 8:21)
dat3<-dat2[myindivs]
#rename myhack to plot
names(dat3)[1]="plot"
```

##What's that look like?
```{r}
kable(head(dat3))
```

##Column headers are values, not variable names
Time for tidyr!: **gather** and **spread**
```{r}
library(tidyr)
```

##Gather the willow individuals by plot and variable
```{r, warning=FALSE}
dat4<-gather(dat3, individ, value, w_1:w_C)
kable(head(dat4))
#unique(dat4$value)
```

##Then spread by variable and value
```{r}
dat5<-spread(dat4, variable, value)
kable(head(dat5))
```

##Uh oh, looks like people treated the ht1 as a notes column!!
What issue is this?
```{r}
unique(dat5$ht1)
```

##Let's fix it
```{r}
dat5<-dat5%>%
  #let's paste the ht1 character values with the notes
  mutate(notes=ifelse(ht1=="CF", paste(notes, "CF", sep=""), notes), 
         notes=ifelse(ht1=="dead", paste(notes, "dead", sep=""), notes)) %>%
  
  #then remove the character values from ht1
  mutate(ht1=ifelse((ht1=="CF" | ht1=="dead"), NA, ht1)) %>%
  
  #but it still thinks it's a factor! fix it to numeric 
  mutate(ht1=as.numeric(as.character(ht1)))
  
```

##What are the other variables?
```{r}
str(dat5)
```

##We want  cnpy1, cnpy2  and ht2 to be numeric!
```{r}
dat5<-dat5%>%
  mutate(cnpy1=as.numeric(cnpy1), cnpy2=as.numeric(cnpy2), 
         ht2=as.numeric(ht2))
str(dat5)
```

##Turns out that some seedlings were seeded, and others transplanted, marked by numeric vs alphabetical
What should we do?
```{r}
dat5<-dat5%>%
  mutate(planting=extract_numeric(individ))%>%
  mutate(planting=ifelse(is.na(planting)==T, "transplant", "seedling"))
```

##Let's more formally characterize mortality:
```{r}
dat5<-dat5%>%
  mutate(dead=0, dead=ifelse(is.na(ht1)==T, 1, dead))
```

##And make one dataset just for live/dead
```{r}
survival_dat<-dat5%>%
  select(plot, individ, dead, planting)
```

##What's that look like?
```{r}
kable(head(survival_dat))
```


##Then let's make a second dataset just for growth
```{r}
growth_dat<-dat5%>%
  filter(dead==0)%>%
  select(plot:ht2, planting)
```

##What's that look like?
```{r}
kable(head(growth_dat))
```

##Is it tidy yet?!





